PAPER DRAFT

### 2022-02-14 Papers/Polymethylation_scores/04_background.Rmd
revise intro and discussion

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
---

@@

Maternal prenatal smoking is a public health concern. In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard for smoking measurements, serum cotinine levels, have a short half-life of nine hours in pregnant women [@Dempsey:2002wr]. Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults [@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical sample, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.

Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across genetic ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of the associations between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed genetic ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association with prenatal maternal smoking among primarily European children, 7 CpG sites were also associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. While prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across genetic ancestry is understudied.

After association testing, an important next step is to evaluate DNA methylation as a biomarker of prenatal maternal smoke exposure by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. Biomarkers can consist of single DNA methylation sites or summary meausres across multiple sites [@Philibert2020; @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. One attractive option is a polymethylation score, analogous to a polygenic score. In the polymethylation score, CpG sites are weighted by the strength of their association with maternal smoking from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. One such DNA methylation score calculated from cord blood classified maternal smoking behavior with an area under the curve (AUC) of 0.82 in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior was classified with an AUC of 0.87 [@Ladd-Acosta:2016aa]. Among middle-aged adults, a score calculated using DNA methylation in peripheral blood samples from a single time point could even predict prenatal smoke exposure (30 years previously) with an AUC of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. Ideally, a biomarker for prenatal maternal smoking would be consistent over participant age at DNA methylation sampling and portable across genetic ancestry groups. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age, and ancestry has not been evaluated in the same study.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoking. Before translation to clinical applications, we must understand how and if the signal varies across age, ancestry, and tissue sample. In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites. We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and consistent biomarker of prenatal maternal smoking.

>>>>

---
output:
  word_document: default
  pdf_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
latex_engine: xelatex
---

@@

In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outisde of birth cohorts[@Dempsey:2002wr]. Alternative biomarkers of prenatal maternal smoke exposure could improve research and clinical practice.

DNA methylation is a possible biomarker. Prenatal maternal smoking is associated with DNA methylation signatures in placental, cord blood, and peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb]. Cord blood and placental samples are rarely available outside of birth cohorts. While peripheral blood is a more realistic sample, saliva is easier to collect [] and increasingly used in large epidemiological cohorts []. Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.

Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017]. Own-smoking DNA methylation signatures differ by genetic ancestry [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012; @Liu2019]. Most work on prenatal maternal smoking and child DNA methylation has been conducted in European-ancestry cohorts [@Joubert:2016ue], with some notable exceptions [@Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty]. Some CpG sites associated with prenatal maternal smoking in European-ancestry cohorts have been replicated in non-European-ancestry or multi-ethnic populations [@Tehranifar:2018tb; @Neophytou:2019ty]. Children of both Black and non-Black identifying mothers exhibited similar directions of association between prenatal maternal smoking and peripheral blood DNA methylation at 38 CpG sites [@Xu2021]. However, this analysis stratified by self-reported ethnicity rather than genetic ancestry. Very studies have evaluated the portability of DNA methylation signatures of prenatal maternal smoke exposure across genetic ancestry populations within a single cohort [@Tehranifar:2018tb; @Gonseth2016].

After association testing, DNA methylation-based biomarkers of prenatal maternal smoking can be assessed by evaluating their classification accuracy. Biomarkers range from single DNA methylation sites to multi-site summary measures, such as polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. In a polymethylation score, CpG sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in newborns and young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly in middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu]. DNA methylation signatures of own smoking change post-cessation [@keshawarz2021]. Associations with prenatal maternal smoking change at some CpG sites as children age [@Richmond2015]. Yet few studies have assessed the longitudinal persistence of DNA-methylation biomarker accuracy.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoking. We must understand how and if the signal varies across tissues sample, genetic ancestry, and age. In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori CpG sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15. We then performed ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.

### 2022-02-14 Papers/Polymethylation_scores/06_discussion.Rmd
revise intro and discussion

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in children's saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores across strata of both child age and genetic ancestry. Global methylation and epigenetic clocks were not associated with maternal smoking exposure. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking (AUC \> 0.7). Classification when using polymethylation scores for prenatal maternal smoking was better than when using a single *a priori* CpG site, cg05575921 in the *AHRR* gene.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated the top hit from a previous epigenome wide association analysis [@Joubert:2016ue]. Further, we found evidence of association in several additional hits from epigenome wide association studies of DNA methylation in cord and peripheral blood [@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017).

We advance prenatal smoking - DNA methylation literature by evaluating the persistence of the association between prenatal maternal smoking and DNA methylation as children age and its portability across tissue and ancestry. Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue]. The portability of other risk scores, such as polygenic risk scores, across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020]. In this case, the polymethylation score for prenatal maternal smoking appears to be portable across genetic ancestry groups. Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower, non-significant effect estimates at age 15. The decreased precision may be the result of the reduced sample size of the HIspanic and European genetic ancestry samples. The reduction in effect estimate magnitude could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation of the effect on *AHRR*:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].

Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds performed comparably to classification using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu]. However, previous prenatal maternal smoke exposure biomarkers created using LASSO regression and cord blood from newborns performed better (AUCs ranging from 0.82-0.97) [@Ladd-Acosta:2016aa] A support vector machine approach performed on and peripheral blood from 3-5 year old children also more accurately classified prenatal maternal smoke exposure (AUC=0.87) [@Reese:2017ts]. Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers. Different methods for summarizing across multiple DNA methylation sites may also influence biomarker performance. The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification. DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement and cell-type proportion [@Hüls2020]. Prenatal maternal smoke exposure classification accuracy of polymethylation score was similar when using coefficients from cord blood vs peripheral blood samples from older children. Classification accuracy was not improved by using two time-points of methylation measurement. Classification accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control. Our analysis therefore suggests that cell-type control when both generating (in the epigenome wide association study) and applying coefficients for polymethylation scores may positively influence outcome classification accuracy, although more work comparing coefficients from different populations is still needed.

Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers. The site cg05575921 in the *AHRR* gene is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue]. DNA methylation at cg05575291 in the *AHRR* gene can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020]. However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu]. In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores. The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures. Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.

@@

However, our analysis also has several strengths. We analyzed samples from a large cohort of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017]. While our exposure measurement of prenatal maternal smoking was self-reported, it was assessed prospectively and preceded outcome measurements. We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch. We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores. In sensitivity analyses, we adjusted for other prenatal exposures and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.

>>>>

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed that prenatal maternal smoking was associated with several characterizations of DNA methylation in childrens saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores for prenatal smoke exposure across strata of genetic ancestry and age. Global methylation and epigenetic clocks were not associated with maternal smoking exposure. Polymethylation scores for prenatal smoke exposure improved classification of self-reported prenatal maternal smoking over a base model, while a single *a priori* CpG site, *AHRR*:cg05575921, did not.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from previous epigenome wide association analyses in cord and peripheral blood [@Joubert:2016ue; @wiklund2019]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, similar populations of immune cells are present in both blood and saliva, and saliva samples in our study were primarily composed of immune cells. Additionally, previous work has shown that the majority of CpG sites are similarly methylated in blood and saliva (reviewed in @Langie2017). Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking.

We advance prenatal smoking - DNA methylation literature by evaluating the portability of association between prenatal maternal smoking and DNA methylation across ancestry groups. The portability of polygenic risk scores across genetic ancestries is a complex research area [@Martin2019] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Huls2020]. Our results suggest polymethylation scores for prenatal maternal smoking are portable across genetic ancestry groups. In the samples from age 15, admixed Latin and European ancestry samples had reduced effect estimates. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, differential initiation and reporting of own-smoking could result in more similar DNA methylation profiles of children exposed vs unexposed to prenatal maternal smoking.

We contribute to the understanding of the impact of time-since-exposure on the classification accuracy of DNA methylation biomarkers for prenatal maternal smoking. Our saliva-based classification of 9- and 15-year-olds (AUC=0.78 and 0.77 respectively) was less accurate than previous cord-blood- and peripheral-blood-based classifications in newborns (AUCs ranging from 0.82-0.97) [@Reese:2017ts] and 3-5-year-old children (AUC=0.87) [@Ladd-Acosta:2016aa]. The difference in classification accuracy could reflect lower proportion of immune cells of saliva as compared to blood. However, they could also reflect a diminishing capacity to distinguish exposure status as time since exposure increases and new environmental exposures are encountered. Indeed, our classification was better than peripheral-blood-based classification in 30-year-old adults (AUC=0.72) [@Richmond:2018tu]. Notably, a previous longitudinal analysis of children's blood samples from birth to 17 years of age found that larger changes in DNA methylation occured between 0 and 7 years of age than between 7 and 17 years of age [@Richmond2015]. In that cohort, the association between prenatal smoke exposure and DNA methylation changed between childhood and adolescence at some but not all CpG sites. This may explain why we found that the polymethylation score improved classification accuracy while *AHRR*:cgg05575921 methylation did not. While *AHRR*:cg05575921 hypomethylation is a consistent marker of prenatal maternal smoking in newborns and children, the association did not persist in samples from adult women [@Richmond:2018tu]. Similarly, while *AHRR*:cg05575291 methylation measured in both blood and saliva can very accurately classify own-smoking [@Dawes2019; @Philibert2020], the magnitude of hypomethylation is reduced post-cessation[@keshawarz2021]. As the accuracy of *AHRR*:cg05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures, incorporating information across multiple sites of DNA methylation may yield a a more robust biomarker.

We note that while the pediatric DNA methylation clock accurately recapitulated subjects chronological ages, the GRIM-age values in our population were overestimated by 10-15 years. In an analysis of 4 pediatric cohorts of children under 20 years of age, GRIM age estimates were 30-35 years greater than chronological in buccal cell samples and 10-15 years greater than chronological age in blood samples [@okazaki2021]. GRIM-age was neither trained nor validated in a pediatric population [@lu2019], which may explain this overestimation.

@@

However, our analysis also has several strengths. We analyzed samples from a large cohort of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017]. While our exposure measurement of prenatal maternal smoking was self-reported, it was assessed prospectively and preceded outcome measurements. We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch. We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores. We adjusted for prenatal maternal drinking, other prenatal drug use and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.
